Recursive Solutions to First-Order Model Counting

### 2022-08-07 doc/paper/paper.tex
Feedback

\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in first-order logic. Despite being around for
  more than a decade, practical FOMC algorithms are still unable to compute
  functions as simple as a factorial. We argue that the capabilities of FOMC
  algorithms are severely limited by their inability to express arbitrary
  recursive computations. To enable arbitrary recursion, we relax the
  restrictions that typically accompany domain recursion and generalise circuits
  used to express a solution to an FOMC problem to graphs that may contain
  cycles. To this end, we enhance the most well-established (weighted) FOMC
  algorithm \textsc{ForcLift} with new compilation rules and an algorithm to
  check whether a recursive call is feasible. These improvements allow us to
  find efficient solutions to counting fundamental structures such as injections
  and bijections.
\end{abstract}

@@

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the size of the instance. The
equivalent notion in (W)FOMC is liftability. A (W)FOMC instance is
\emph{(domain-)liftable} if it can be solved in time polynomial in the size(s)
of the domain(s) \citep{jaeger2012liftability}. Over more than a decade, many
classes of instances were shown to be liftable
\citep{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, \citet{DBLP:conf/nips/Broeck11} showed that the class of all sentences of
FOL with up to two variables (denoted \FOtwo{}) is liftable. Then
\citet{DBLP:conf/pods/BeameBGS15} proved that there exists a sentence with three
variables for which FOMC is $\#\P_{1}$-complete (i.e., \FOthree{} is not
liftable). Since these two results came out, most of the research on (W)FOMC
focused on developing faster solutions for the \FOtwo{} fragment
\citep{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining new
liftable fragments. These fragments include \SFO{} and \SRU{}
\citep{DBLP:conf/nips/KazemiKBP16}, \Uone{} \citep{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \citep{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the
empirical front, there are several open-source implementations of exact WFOMC
algorithms: \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving \citep{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C} \citep{DBLP:conf/kr/KazemiP16}. Approximate counting is supported
by \textsc{ApproxWFOMC} \citep{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \citep{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \citep{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
that is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains are added to $\mathcal{D}$ by some of
the compilation rules.

@@

One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

% \footnote{If some domain is not big enough to contain all of its constants,
% the formula is unsatisfiable.}

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.\footnote{If $(X, Y)$ is a two-variable constraint, substituting a
  constant $c$ for $X$ would result in $(c, Y)$, which would have to be
  rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

@@

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!10}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!10}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!10}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!10}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & --- & $mn$ \\
    \rowcolor{gray!10}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions.
    Here, $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that no complete solution was found.}\label{tbl:results}
\end{table*}

@@

>>>>

% TODO: FOMC algorithms -> of most algorithms to date
\begin{abstract}
  First-order model counting (FOMC) is a computational problem that asks to
  count the models of a sentence in finite-domain first-order logic. Despite
  being around for more than a decade, practical FOMC algorithms are still
  unable to compute functions as simple as a factorial. We argue that the
  capabilities of FOMC algorithms are severely limited by their inability to
  express arbitrary recursive computations. To enable arbitrary recursion, we
  relax the restrictions that typically accompany domain recursion and
  generalise circuits used to express a solution to an FOMC problem to graphs
  that may contain cycles. To this end, we enhance the most well-established
  (weighted) FOMC algorithm \textsc{ForcLift} with new compilation rules and an
  algorithm to check whether a recursive call is feasible. These improvements
  allow us to find efficient solutions to counting fundamental structures such
  as injections and bijections.
\end{abstract}
% TODO: We conclude after report on empirical results and discuss a few
% conjectures on what correctness might look like.

@@

% TODO: It might be worth remarking either in the abstract or in the intro that
% generally the weighted version is what is studied but the nonweighted version
% is also still interesting because of its connections to finite model theory.
% You could make a reference to the work of Paul Beam on symmetric weighted
% model counting.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\citep{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be liftable
\citep{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, \citet{DBLP:conf/nips/Broeck11} showed that the class of all sentences of
FOL with up to two variables (denoted \FOtwo{}) is liftable. Then
\citet{DBLP:conf/pods/BeameBGS15} proved that there exists a sentence with three
variables for which FOMC is $\#\P_{1}$-complete (i.e., \FOthree{} is not
liftable). Since these two results came out, most of the research on (W)FOMC
focused on developing faster solutions for the \FOtwo{} fragment
\citep{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining new
liftable fragments. These fragments include \SFO{} and \SRU{}
\citep{DBLP:conf/nips/KazemiKBP16}, \Uone{} \citep{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \citep{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\citep{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the
empirical front, there are several open-source implementations of exact WFOMC
algorithms: \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving \citep{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C} \citep{DBLP:conf/kr/KazemiP16}. Approximate counting is supported
by \textsc{ApproxWFOMC} \citep{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \citep{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \citep{DBLP:journals/cacm/GogateD16}.

@@

% TODO (footnote): You need to explain this a little bit more. You say that it
% is liftable in principle but you should clarif ythat factorial has never been
% shown empirically and there is no correctness theorem claiming that a
% factorial can be computed in a lifted manner. Or something like that.
However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
that is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.
% TODO: Because this is the focus of your work you might want to give at least
% one example of such a stringent condition

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. A \emph{term} is either a constant or a variable. A
\emph{literal} is either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains are added to $\mathcal{D}$ by some of
the compilation rules.
% TODO ('Let D... all domains'): You need to state explicitly or implicitly that
% to set of domains is constructed from a finite set of constants. Either you
% can define the set of constants in the langoage or you can say that in the
% interest of simplicity the set of constants is a set of nullary functions
% mentioned in the sentence, which only mentions relations or nullary functions.
% Function that take arguments are not allowed.

@@

% TODO: You should mention that somewhat unusually, first order variables are in
% capital letters, similar to the convention of logic programming. They are not
% to be confused with second order qualification.
One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}
% TODO: It might be useful to give one or two readable examples with interesting
% model counts. For example Px v Qx over domain n has 3^n/4^n solutions. I think
% The Frends smokers example has a more complicated solution you can find in one
% of the papers. It might motivate the framework generally speaking better
% although of course you can continue to use the same running example for the
% rest of your paper. Such examples are especially useful for readers who are
% not familiar but still need to get excited by the problem

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.

@@

% TODO: Again for the reader who is not very familiar with this area it might be
% good to at least give the circuit for one of the examples such as a
% disjunction just to show them how it is usually done.

@@

% TODO: I think there is a point of confusion here. Firstly the work of guy
% introduce circuits and I presume that you want to say that the node types are
% present in both the circuits as well as your structure. Given that I think you
% should start of the paragraph by saying that like with circuit introduced by
% Guy, there are node type such as XYZN. In addition to those you will be
% introducing the following new ones.
%
% I think the other point that is not made very clear is that you have no
% formally defined circuits. So perhaps if you could introduce a quick
% definition and then say that we have the same building blocks as those, ie
% nodes, but with a different graphical structure it will become clearer

@@

% TODO: explain notation (i.e., remove the bot)
\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2')$, where
  \begin{align*}
    c_1' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!10}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!10}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!10}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!10}\cmark & \xmark & & $\min\{\, m, n \,\}^2$ & --- & $mn$ \\
    \rowcolor{gray!10}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions.
    Here, $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that no complete solution was found.}\label{tbl:results}
\end{table*}
% TODO: This is very neat but readers from the community will expect to see some
% plots of time. Can you also give a few of those that reflect or mirror the
% conclusions from the table? Also it is a standard to discuss a few lifted
% theories like the smokers and friends example. Can you also include some
% empirical results with plots of time on such theories as well to appeal to the
% sanity of the reader that it actually works for problems in frameworks such
% as Markov log networks

@@

% TODO: You previously mentioned this class being discussed in another paper.
% You will need to remind the reader what you are doing differently to that and
% how you are improving to previous work.
